Early Warning Signals and the Prosecutor's Fallacy

### 2012-05-30 fallacy.tex
edits

  Early warning signals have been proposed to forecast the possibility of a 
  critical transition, such as the eutrophication of a lake, desertification
  of savanna, the collapse of a coral reef, the end of a glacial period, or 
  the onset of epilepsy in the brain.  While such transitions may be difficult
  to study in the laboratory setting and unethical to introduce on the larger
  natural scales where they may occur, it has often been observed that history
  has already performed the experiment for us, time and again.  Here we point 
  out a critical difference between selecting systems for study based on the 
  fact that we have observed a critical transition and those systems for which
  we wish to forecast the approach of a transition. This difference arises
  by conditionally selecting systems known to experience a transition of some 
  sort and failing to account for this selection step -- a statistical error 
  often known as the Prosecutor's
  Fallacy.  We illustrate through simulation how examples of systems that have 
  experienced transitions purely by chance show patterns associated with 
  early warning signals more often than we might expect in systems that have not
  been conditionally selected.  We further highlight how a model-based approach
  is less subject to this bias.  In using historical examples of collapse to 
  test and validate methods for early warning signals, greater care must be taken
  to avoid this fallacy.  Groundbreaking experimental studies with replicates
  avoid this pitfall, and more of such studies are needed to provide data
  suitable for evaluating methods for early warning.

@@

\subsection*{Catastrophic Transitions}

Catastrophic transitions or tipping points, where a complex system 
shifts suddenly from one state to another, have been implicated in
a wide array of ecological and global climate systems such as lake
ecosystems~\citep{Carpenter2011}, coral reefs~\citep{Mumby2007},
desertification~\citep{Kefi2007}, fisheries~\citep{Berkes2006}, and 
tropical forests~\citep{Hirota2011}.  Recent research has begun to
identify statistical patterns commonly associated with these sudden
catastrophic transitions which could be used as an \emph{early warning sign}
to identify an approaching tipping point, buying precious time 
to react and avert them~\citep{Scheffer2009, Lenton2011}.

An array of statistical patterns associated with tipping point phenomena has
been suggested for the detection of early warning signals associated with
such sudden transitions.  Two of the most commonly used are a pattern of
increasing variance~\citep{Carpenter2006} and a pattern of increasing 
autocorrelation~\citep{vanNes2007}, which have been tested in both experimental
manipulation~\citep{Drake2010, Carpenter2011, Veraart2011} and historical 
observations~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}

@@

Historical examples of sudden transitions taken from the paleoclimate record
provide an important way to test and evaluate potential leading indicator 
methods, and have been widely used for this purpose 
\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}.
Similarly, it has been suggested that data gathered from ecological systems such
as lakes that were monitored before they collapsed, or grasslands subjected to
overgrazing, could contain data that could help reveal when similar systems 
are approaching a tipping point~\citep{Inman2011}.  This is a silver lining to
the clouds of past catastrophes -- we may be able to learn from our mistakes.

Yet in testing methods for early warning signals against historical examples of
transitions, we must be careful to avoid statistical mistakes that arise when
we select examples conditional on having exhibited a sudden transition.  Such
sudden transitions can occur for many reasons that do not involve the 
bifurcation processes on which most early warning indicators are predicated
~\citep{Hastings2010}, including purely noise-induced transitions for proposed 
early warning patterns are not anticipated~\citep{Ditlevsen2010, Lenton2011}. 
Ignoring the potential biases introduced by conditionally selecting from 
historical records is an example of the statistical error commonly known as the
\emph{Prosecutor's Fallacy.}

\subsection*{The Prosecutor's Fallacy}

In a criminal trial, the prosecution may introduce associative evidence such as 
blood testing to establish a match between the perpetrator at the scene of the
crime and a suspect on the defense stand.  Statistical testimony may accompany 
such evidence reflecting the incidence rate of the matching characteristic -- 
for instance, the frequency of individuals in the population with the matching
blood type.  Such arguments may be misleading when they fail to account for  
the manner in which the suspect was identified in 
the first place.  If the suspect has been selected partly on the basis of 
belonging to a demographic in which the discriminating characteristic is much more common than
in the population as a whole, these odds can be misleading.  Failing to recognize 
these conditional probabilities leads to the Prosecutor's Fallacy: the 
argument that the suspect is the perpetrator on the basis of the statistical 
probability of a match without correctly identifying the baseline probability 
for such a match \citep{Thompson1987}.

@@

\section{Methods and Results}
We approach the problem by simulating a system that cannot undergo the
saddle node bifurcation that underlies the usual early warning signal
indicators.  This simulation approach is the only way to determine whether
examining historical events is a good way to test the utility of these
indicators.  We simulated 1000 replicates of a stochastic individual-based
birth-death process with an Allee threshold~\citep{Courchamp2008}, which
arises from positive fitness effects at low densities.  Above the Allee
threshold the population returns to a positive equilibrium size, while
below the threshold the population decreases to zero in expectation. The
simulation uses the Gillespie algorithm to provide an exact implementation
of the master equation of the birth-death process,

@@

To test for the presence of a warning signal in these indicators we
computed values of Kendall's $\tau$ for both indicators for each of
the 266 replicates.  Kendall's $tau$ takes values in $(-1, 1)$, and is
frequently used to identify an increasing trend ($\tau > 0 )$ in early
warning signals.~\citep{Dakos2008, Dakos2011}.  The distribution
of $\tau$ values observed across these replicates is shown in
Figure~\ref{fig:indicator}.  More positive values indicate stronger
trends. The common statistical significance test used for Kendall's $tau$
assumes independence of the variables to achieve a normal distribution,
this isn't satisfied for indicators coming from a time series and
computed with a sliding window.  For a more robust comparison, we
compare the distribution of this statistic from all the simulations to
the distribution conditioned on experiencing a chance transition to the
alternative stable state, that is, from the replicates which are known
to crash.  To avoid an effect of sample size the time series are all chosen
to be the same length.  As these examples have been selected from a system
that is not approaching a bifurcation, an indicator that does not suffer
from the bias of Prosecutor's fallacy should have the same distribution
in both cases.   Annotated, parallelized code to replicate the analysis
and produce the figures is provided in the appendix.  (Currently at
\href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).

@@

 \section{Acknowledgements}
This research was supported by funding from NSF Grant EF 0742674 to AH
and a Computational Sciences Graduate Fellowship from the Department of
Energy grant DE-FG02-97ER25308 to CB.

>>>>

  Early warning signals have been proposed to forecast the possibility
  of a critical transition, such as the eutrophication of a lake,
  the collapse of a coral reef, or the end
  of a glacial period.  Because such transitions often occur on scales that
  can be difficult to approach by experimental manipulation, research
  has often relied on historical observations as a source of natural experiments.
  Here we examine a critical difference between selecting systems for
  study based on the fact that we have observed a critical transition
  and those systems for which we wish to forecast the approach of a
  transition. This difference arises by conditionally selecting systems
  known to experience a transition of some sort and failing to account
  for this bias this introduces -- a statistical error often known as the
  Prosecutor's Fallacy.  By analyzing simulated systems that have
  experienced transitions purely by chance, we reveal an elevated rate
  of false postives in common warning signal statistics. We further
  demonstrate a model-based approach which is less subject to this bias
  than these more commonly used summary statistics.
  In using historical examples of collapse to test and validate methods
  for early warning signals, care must be taken to avoid this fallacy,
  either through the use methods less sensitive to this effect or ideally
  by greater reliance on Experimental studies with replicates avoid this 
  pitfall entirely.  More such studies are needed to provide data suitable
  for evaluating methods for early warning.

@@

%\subsection*{Catastrophic Transitions}

Catastrophic transitions or tipping points, where a complex system
shifts suddenly from one state to another, have been implicated in
a wide array of ecological and global climate systems such as lake
ecosystems~\citep{Carpenter2011}, coral reefs~\citep{Mumby2007},
savannah~\citep{Kefi2007}, fisheries~\citep{Berkes2006}, and tropical
forests~\citep{Hirota2011}.  Recent research has begun to identify
statistical patterns commonly associated with these sudden catastrophic
transitions which could be used as an \emph{early warning sign} to
identify an approaching tipping point, which might provide managers time
to react to and avert an undesirable state shift~\citep{Scheffer2009,
Lenton2011}.  An array of statistical patterns associated with tipping
point phenomena has been suggested for the detection of early warning
signals associated with such sudden transitions.  Two of the most commonly
used are a pattern of increasing variance~\citep{Carpenter2006} and a
pattern of increasing autocorrelation~\citep{vanNes2007}, which have
been tested in both experimental manipulation~\citep{Drake2010,
Carpenter2011, Veraart2011} and historical
observations~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}.

@@

Historical examples of sudden transitions taken from the paleoclimate
record provide an important way to test and evaluate potential
leading indicator methods, and have been widely used for this purpose
\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}.
Similarly, it has been suggested that data gathered from ecological
systems such as lakes that were monitored before they experienced sudden
eutriphication, or grasslands subjected to overgrazing, could contain
data that could help reveal when similar systems are approaching a
tipping point~\citep{Carpenter2011}.  This is a silver lining to the
clouds of past catastrophes; we may be able to learn from such examples 
how to detect and avert sudden shifts in the future.

However, testing methods for early warning signals against historical examples of
transitions is suspetible to statistical mistakes that arise from selecting
data conditional on that data having already exhibited a sudden transition.  
A central tenent of early warning theory is that the system in question is slowly
approaching a tipping point that lies some unknown distance away.  If nothing
is done to remedy the situation, this slow change will inevitably carry the 
system beyond the tipping point, which introduces a sudden, rapid transition into an 
undesirable state~\citep{Scheffer2009}. This process can be described mathematically
as a \emph{bifurcation}, in which a slowly changing parameter reaches a critical
value that causes the system stability to change.

Such
sudden transitions can occur for many reasons that do not involve the 
bifurcation processes on which most early warning indicators are predicated
~\citep{Hastings2010}, including purely noise-induced transitions for proposed 
early warning patterns are not anticipated~\citep{Ditlevsen2010, Lenton2011}. 
Ignoring the potential biases introduced by conditionally selecting from 
historical records is an example of the statistical error commonly known as the
Prosecutor's Fallacy.

In a criminal trial, the prosecution may introduce associative evidence such as 
blood testing to establish a match between the perpetrator at the scene of the
crime and a suspect on the defense stand.  Statistical testimony may accompany 
such evidence reflecting the incidence rate of the matching characteristic,
such as the frequency of individuals in the population with the matching
blood type.  Such arguments are misleading when they fail to account for  
the manner in which the suspect was identified in the first place.  If the suspect has been selected on the basis of 
belonging to a demographic in which the discriminating characteristic is much more common than
in the population as a whole, .

Failing to recognize 
these conditional probabilities leads to the Prosecutor's Fallacy: the 
argument that the suspect is the perpetrator on the basis of the statistical 
probability of a match without correctly identifying the baseline probability 
for such a match \citep{Thompson1987}.

@@

\section{Methods and Results}
We approach the problem by simulating a system that will not undergo the
saddle node bifurcation that underlies the usual early warning signal
indicators.  This simulation approach allows us to determine whether
examining historical events is a good way to test the utility of these
indicators.  We simulated 1000 replicates of a stochastic individual-based
birth-death process with an Allee threshold~\citep{Courchamp2008}, which
arises from positive fitness effects at low densities.  Above the Allee
threshold the population returns to a positive equilibrium size, while
below the threshold the population decreases to zero in expectation. The
simulation uses the Gillespie algorithm to provide an exact implementation
of the master equation of the birth-death process,

@@

To test for the presence of a warning signal in these indicators we
computed values of Kendall's $\tau$ for both indicators for each of
the 266 replicates.  Kendall's $\tau$ takes values in $(-1, 1)$, and is
frequently used to identify an increasing trend ($\tau > 0 )$ in early
warning signals.~\citep{Dakos2008, Dakos2011}.  The distribution
of $\tau$ values observed across these replicates is shown in
Figure~\ref{fig:indicator}.  More positive values indicate stronger
trends. The common statistical significance test used for Kendall's $\tau$
assumes independence of the variables to achieve a normal distribution,
this isn't satisfied for indicators coming from a time series and
computed with a sliding window.  For a more robust comparison, we
compare the distribution of this statistic from all the simulations to
the distribution conditioned on experiencing a chance transition to the
alternative stable state, that is, from the replicates which are known
to crash.  To avoid an effect of sample size the time series are all chosen
to be the same length.  As these examples have been selected from a system
that is not approaching a bifurcation, an indicator that does not suffer
from the bias of Prosecutor's fallacy should have the same distribution
in both cases.   Annotated, parallelized code to replicate the analysis
and produce the figures is provided in the appendix.  (Currently at
\href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).

@@

 \section{Acknowledgements}
This research was supported by funding from NSF Grant EF 0742674 to AH
and a Computational Sciences Graduate Fellowship from the Department of
Energy grant DE-FG02-97ER25308 to CB. The authors thank M. Baskett, 
T.A. Perkins and N. Ross for helpful comments on earlier drafts of the
manuscript.
